---
title: "Notebook 1 — R Reboot for Bioinformatics"
format: html
editor: visual
---

# Goal

Rebuild core R fluency for bioinformatics using short, hands-on examples.  
This notebook focuses on variables, strings, vectors, indexing, functions, lists, and simple data frames.

## Section 1 — Running R Code

Concept: variables and assignment.

```{r}
sequence <- "ATGCGT"
print(sequence)
```

### Exercises

1. Store a gene name in a variable.
2. Store a DNA sequence.
3. Print both variables.

---

## Section 2 — Strings

Concepts:

- `nchar()`
- `substr()`
- `toupper()`

```{r}
seq <- "ATGCGT"

nchar(seq)
substr(seq, 1, 3)
toupper(seq)
```

### Exercises

4. Print sequence length.  
5. Extract the first three bases.  
6. Extract the last three bases.

---

## Section 3 — Vectors

R revolves around vectors.

```{r}
genes <- c("TP53", "KRAS", "BRAF")

genes
length(genes)
```

### Exercises

7. Create a vector of gene names.  
8. Add another gene to the vector.  
9. Print the second gene.

---

## Section 4 — Vectorized Operations

R operates on vectors directly.

```{r}
bases <- c("A", "T", "G", "C", "G")

bases == "G"
sum(bases == "G")
```

### Exercises

10. Identify all `"C"` bases.  
11. Count `"G"` bases.

---

## Section 5 — Logical Filtering

```{r}
genes <- c("TP53", "KRAS", "BRAF", "KRAS")

genes[genes == "KRAS"]
```

### Exercises

12. Extract all `"KRAS"` entries.  
13. Count KRAS entries.

---

## Section 6 — Simple Loops

Loops exist in R, though vectorized operations are often preferred.

```{r}
seq_vec <- c("A", "T", "G", "C")

for (b in seq_vec) {
  print(b)
}
```

### Exercises

14. Loop through a sequence and print each base.

---

## Section 7 — Basic Functions

```{r}
gc_content <- function(seq) {
  bases <- strsplit(seq, "")[[1]]
  gc <- sum(bases %in% c("G", "C"))
  gc / length(bases)
}

gc_content("ATGCGT")
```

### Exercises

15. Write a function returning sequence length.  
16. Write a function returning GC count.

---

## Section 8 — Lists

Lists can store mixed objects.

```{r}
sample_mutation <- list(
  gene = "TP53",
  position = 273,
  mutation = "R273H"
)

sample_mutation$gene
sample_mutation$mutation
```

### Exercises

17. Create a list representing a mutation.  
18. Extract the mutation field.

---

## Section 9 — Simple Data Frames

```{r}
mutations <- data.frame(
  patient = c("P1", "P2", "P3"),
  gene = c("TP53", "KRAS", "BRAF"),
  mutation = c("R273H", "G12D", "V600E")
)

mutations
```

### Exercises

19. Print the gene column.  
20. Print the first two rows.

---

## Section 10 — Simple Summaries

```{r}
table(mutations$gene)
table(mutations$patient)
```

### Exercises

21. Count mutations per gene.  
22. Count mutations per patient.

---

## Section 11 — Mini Bio Problem

Compute GC content for multiple sequences.

```{r}
sequences <- c("ATGC", "GGGG", "TTTT")

sapply(sequences, gc_content)
```

### Exercises

23. Compute GC content for each sequence.  
24. Find sequences with GC > 0.5.  
25. Print their indices.

---

## Skills Gained After Notebook 1

By the end of this notebook, you should be comfortable with:

- variables and assignment
- strings and indexing
- vectors and vectorized operations
- logical filtering
- simple loops
- functions
- lists
- simple data frames
- summary tables

## Next Step

Notebook 2 will move into structured data manipulation in R, focusing on:

- data frames
- filtering rows
- summarizing data
- reshaping tables
- biological example datasets
