---
title: "Notebook 2 — Data Frames and Structured Data in R"
format: html
editor: visual
---

# Goal

Learn how to work with structured biological datasets in R using **data frames**.
This notebook introduces filtering, summarizing, and basic manipulation of tabular data.

---

## Section 1 — Creating Data Frames

Data frames are the core structure for tabular biological data in R.

```{r}
mutations <- data.frame(
  patient = c("P1", "P2", "P3", "P4", "P5"),
  gene = c("TP53", "KRAS", "TP53", "BRAF", "KRAS"),
  mutation = c("R273H", "G12D", "R175H", "V600E", "G12V")
)

mutations
```

### Exercises

1. Print the data frame.
2. Print only the first three rows.

---

## Section 2 — Accessing Columns

Columns can be accessed using `$` or indexing.

```{r}
mutations$gene
mutations[["mutation"]]
```

### Exercises

3. Print the patient column.
4. Print both the gene and mutation columns.

---

## Section 3 — Row Subsetting

We can filter rows using logical conditions.

```{r}
mutations[mutations$gene == "KRAS", ]
```

### Exercises

5. Extract rows where gene is `"TP53"`.
6. Extract rows where gene is `"BRAF"`.

---

## Section 4 — Logical Conditions

Multiple conditions can be combined.

```{r}
mutations[mutations$gene == "KRAS" | mutations$gene == "TP53", ]
```

### Exercises

7. Extract rows where gene is KRAS or BRAF.
8. Extract rows where patient is P1 or P2.

---

## Section 5 — Basic Summaries

R has simple built‑in functions for summarizing categorical data.

```{r}
table(mutations$gene)
```

### Exercises

9. Count mutations per gene.
10. Count mutations per patient.

---

## Section 6 — Adding Columns

New columns can be added to a data frame.

```{r}
mutations$gene_length <- nchar(mutations$gene)

mutations
```

### Exercises

11. Add a column with mutation string length.
12. Add a column indicating whether the gene is KRAS.

---

## Section 7 — Ordering Data

Rows can be sorted.

```{r}
mutations[order(mutations$gene), ]
```

### Exercises

13. Sort rows by patient.
14. Sort rows by mutation.

---

## Section 8 — Unique Values

Identify unique values in a column.

```{r}
unique(mutations$gene)
```

### Exercises

15. Find unique patients.
16. Count how many unique genes appear.

---

## Section 9 — Basic Aggregation

Aggregate functions summarize data by groups.

```{r}
aggregate(patient ~ gene, data = mutations, FUN = length)
```

### Exercises

17. Count mutations per gene using `aggregate()`.
18. Compute the number of mutations per patient.

---

## Section 10 — Mini Bioinformatics Dataset

Create a small dataset representing gene expression values.

```{r}
expression <- data.frame(
  gene = c("TP53", "KRAS", "BRAF", "EGFR"),
  sample1 = c(10, 50, 30, 40),
  sample2 = c(12, 60, 35, 42)
)

expression
```

### Exercises

19. Print the expression matrix.
20. Extract expression values for KRAS.

---

## Section 11 — Mini Analysis

Compute the average expression per gene.

```{r}
expression$mean_expression <- rowMeans(expression[,2:3])

expression
```

### Exercises

21. Compute mean expression per gene.
22. Identify the gene with the highest expression.

---

## Section 12 — Simple Data Exploration

Inspect the structure of a dataset.

```{r}
str(mutations)
summary(mutations)
```

### Exercises

23. Inspect the structure of the expression dataset.
24. Print summary statistics.

---

## Final Exercise

25. Build a small analysis that:

- counts mutations per gene
- sorts the results
- prints the most frequently mutated gene

---

## Skills Gained After Notebook 2

You should now understand:

- how to create data frames
- column and row indexing
- filtering rows with logical conditions
- counting categorical data
- adding columns
- sorting rows
- summarizing tabular datasets

---

## Next Step

Notebook 3 will introduce the **tidyverse workflow**, including:

- `dplyr`
- pipes
- grouped summaries
- cleaner data manipulation
