---
title: "Notebook 3 — Tidyverse Workflows for Bioinformatics"
format: html
editor: visual
---

# Goal

Introduce the **tidyverse workflow** for analyzing biological datasets in R.
This notebook focuses on `dplyr`, the pipe operator, and grouped summaries.

The tidyverse provides a powerful and readable way to manipulate tabular data.

---

## Section 1 — Loading the Tidyverse

```{r}
library(dplyr)
```

The **pipe operator** (`|>`) allows us to write data transformations in a readable sequence.

---

## Section 2 — Example Mutation Dataset

```{r}
mutations <- data.frame(
  patient = c("P1","P2","P3","P4","P5","P6"),
  gene = c("TP53","KRAS","TP53","BRAF","KRAS","TP53"),
  mutation = c("R273H","G12D","R175H","V600E","G12V","R248Q")
)

mutations
```

### Exercises

1. Print the dataset.
2. Count the number of rows.

---

## Section 3 — Selecting Columns

```{r}
mutations |> select(gene, mutation)
```

### Exercises

3. Select only the patient column.
4. Select both patient and gene columns.

---

## Section 4 — Filtering Rows

```{r}
mutations |> filter(gene == "KRAS")
```

### Exercises

5. Filter rows where gene is TP53.
6. Filter rows where patient is P1 or P2.

---

## Section 5 — Creating New Columns

```{r}
mutations |> 
  mutate(gene_length = nchar(gene))
```

### Exercises

7. Create a column storing mutation string length.
8. Create a column indicating whether gene == "KRAS".

---

## Section 6 — Counting Observations

```{r}
mutations |> count(gene)
```

### Exercises

9. Count mutations per gene.
10. Count mutations per patient.

---

## Section 7 — Grouped Summaries

```{r}
mutations |>
  group_by(gene) |>
  summarise(count = n())
```

### Exercises

11. Count mutations per gene.
12. Count mutations per patient.

---

## Section 8 — Sorting Results

```{r}
mutations |>
  count(gene) |>
  arrange(desc(n))
```

### Exercises

13. Sort genes by mutation frequency.
14. Identify the most frequently mutated gene.

---

## Section 9 — Example Expression Dataset

```{r}
expression <- data.frame(
  gene = c("TP53","KRAS","BRAF","EGFR"),
  sample1 = c(10,50,30,40),
  sample2 = c(12,60,35,42),
  sample3 = c(11,55,32,39)
)

expression
```

### Exercises

15. Print the dataset.
16. Extract the KRAS row.

---

## Section 10 — Row-wise Summaries

```{r}
expression |> 
  mutate(mean_expression = rowMeans(across(sample1:sample3)))
```

### Exercises

17. Compute mean expression per gene.
18. Identify the gene with the highest expression.

---

## Section 11 — Pivoting Data

Reshaping data is common in bioinformatics workflows.

```{r}
library(tidyr)

expression_long <- expression |>
  pivot_longer(cols = sample1:sample3,
               names_to = "sample",
               values_to = "expression")

expression_long
```

### Exercises

19. Inspect the long format dataset.
20. Count rows in the long dataset.

---

## Section 12 — Grouped Analysis on Long Data

```{r}
expression_long |>
  group_by(gene) |>
  summarise(mean_expression = mean(expression))
```

### Exercises

21. Compute mean expression per gene.
22. Compute mean expression per sample.

---

## Final Exercise

23–25. Build a small analysis pipeline that:

- reshapes the expression dataset
- computes mean expression per gene
- sorts genes by expression level

---

## Skills Gained After Notebook 3

You should now understand:

- the tidyverse workflow
- pipes (`|>`)
- `select()`
- `filter()`
- `mutate()`
- `count()`
- `group_by()` and `summarise()`
- reshaping data with `pivot_longer()`

---

## Next Step

Notebook 4 will introduce **visualization using ggplot2**, including:

- bar plots
- scatter plots
- boxplots
- faceting
