---
title: "Notebook 4 — Visualization with ggplot2 for Bioinformatics"
format: html
editor: visual
---

# Goal

Learn how to create clear and informative visualizations for biological datasets using **ggplot2**.

This notebook introduces the **grammar of graphics** approach used by ggplot2, which builds plots in layers.

---

## Section 1 — Load Libraries

```{r}
library(ggplot2)
library(dplyr)
```

---

## Section 2 — Example Mutation Dataset

```{r}
mutations <- data.frame(
  patient = c("P1","P2","P3","P4","P5","P6"),
  gene = c("TP53","KRAS","TP53","BRAF","KRAS","TP53"),
  mutation = c("R273H","G12D","R175H","V600E","G12V","R248Q")
)

mutations
```

### Exercises

1. Print the dataset.
2. Count mutations per gene.

---

## Section 3 — Basic Bar Plot

Bar plots are useful for visualizing categorical counts.

```{r}
ggplot(mutations, aes(x = gene)) +
  geom_bar()
```

### Exercises

3. Create a bar plot of mutation counts by gene.
4. Try reordering the genes using `arrange()` before plotting.

---

## Section 4 — Counting with dplyr then Plotting

```{r}
gene_counts <- mutations |>
  count(gene)

gene_counts
```

```{r}
ggplot(gene_counts, aes(x = gene, y = n)) +
  geom_col()
```

### Exercises

5. Plot mutation counts per gene.
6. Sort the dataset before plotting.

---

## Section 5 — Scatter Plot Example

Example dataset representing biological features.

```{r}
features <- data.frame(
  sample = paste0("S",1:8),
  gc_content = c(0.40,0.55,0.62,0.35,0.70,0.48,0.66,0.30),
  length = c(100,120,150,90,180,110,160,80)
)

features
```

```{r}
ggplot(features, aes(x = gc_content, y = length)) +
  geom_point()
```

### Exercises

7. Plot GC content vs sequence length.
8. Identify any visible patterns.

---

## Section 6 — Boxplots

Boxplots help visualize distributions.

```{r}
features$group <- c("A","A","A","A","B","B","B","B")

ggplot(features, aes(x = group, y = length)) +
  geom_boxplot()
```

### Exercises

9. Plot sequence length distributions by group.
10. Compare the median values visually.

---

## Section 7 — Adding Labels

```{r}
ggplot(features, aes(x = gc_content, y = length)) +
  geom_point() +
  labs(
    title = "GC Content vs Sequence Length",
    x = "GC Content",
    y = "Sequence Length"
  )
```

### Exercises

11. Add a plot title.
12. Change axis labels.

---

## Section 8 — Faceting

Faceting creates multiple panels based on a categorical variable.

```{r}
ggplot(features, aes(x = gc_content, y = length)) +
  geom_point() +
  facet_wrap(~group)
```

### Exercises

13. Create separate panels by group.
14. Compare patterns across groups.

---

## Section 9 — Themes

Themes change plot appearance.

```{r}
ggplot(features, aes(x = gc_content, y = length)) +
  geom_point() +
  theme_minimal()
```

### Exercises

15. Try `theme_bw()`.
16. Compare visual styles.

---

## Section 10 — Visualizing Expression Data

```{r}
expression <- data.frame(
  gene = c("TP53","KRAS","BRAF","EGFR"),
  expression = c(10,50,30,40)
)

ggplot(expression, aes(x = gene, y = expression)) +
  geom_col()
```

### Exercises

17. Plot gene expression values.
18. Identify the highest expressed gene.

---

## Section 11 — Combining dplyr and ggplot

```{r}
mutations |>
  count(gene) |>
  ggplot(aes(x = gene, y = n)) +
  geom_col()
```

### Exercises

19. Create a pipeline that counts mutations and plots them.
20. Sort the plot by mutation frequency.

---

## Final Exercise

21–25. Build a visualization workflow that:

- summarizes a dataset
- creates at least two plots
- compares two biological variables
- labels the plots clearly
- uses a clean theme

---

## Skills Gained After Notebook 4

You should now understand:

- how ggplot2 builds plots in layers
- bar plots for categorical data
- scatter plots for relationships
- boxplots for distributions
- faceting for grouped comparisons
- labeling and themes
- integrating `dplyr` pipelines with plotting

---

## Next Step

Notebook 5 will introduce **statistical analysis and modeling in R**, including:

- linear models
- logistic regression
- PCA
- clustering
