---
title: "Notebook 5 — Statistics and Modeling in R for Bioinformatics"
format: html
editor: visual
---

# Goal

Introduce basic **statistical modeling workflows in R** using small biological datasets.

This notebook covers:

- linear models
- logistic regression
- principal component analysis (PCA)
- clustering

These tools form the foundation of many statistical analyses in computational biology.

---

## Section 1 — Example Dataset

```{r}
dataset <- data.frame(
  sample = paste0("S",1:10),
  gc_content = c(0.40,0.55,0.62,0.35,0.70,0.48,0.66,0.30,0.59,0.45),
  length = c(100,120,150,90,180,110,160,80,140,105),
  motif_count = c(1,3,4,0,5,2,4,0,3,1),
  responder = c(0,1,1,0,1,0,1,0,1,0)
)

dataset
```

### Exercises

1. Print the dataset.
2. Count the number of samples.

---

## Section 2 — Linear Models

Linear models estimate relationships between variables.

```{r}
model <- lm(length ~ gc_content, data = dataset)

summary(model)
```

### Exercises

3. Fit a linear model predicting length from motif_count.
4. Interpret the slope coefficient.

---

## Section 3 — Logistic Regression

Logistic regression predicts binary outcomes.

```{r}
log_model <- glm(responder ~ gc_content + motif_count,
                 data = dataset,
                 family = binomial)

summary(log_model)
```

### Exercises

5. Identify which predictors influence response.
6. Interpret the sign of coefficients.

---

## Section 4 — Model Predictions

```{r}
dataset$predicted_prob <- predict(log_model, type = "response")

dataset
```

### Exercises

7. Print predicted probabilities.
8. Identify samples with probability > 0.5.

---

## Section 5 — Principal Component Analysis

PCA reduces dimensionality of data.

```{r}
pca <- prcomp(dataset[,2:4], scale = TRUE)

summary(pca)
```

### Exercises

9. Inspect variance explained by PC1 and PC2.
10. Identify which variables contribute most.

---

## Section 6 — PCA Visualization

```{r}
pca_scores <- data.frame(pca$x)

pca_scores
```

### Exercises

11. Inspect the PCA coordinates.
12. Identify samples with similar profiles.

---

## Section 7 — Clustering

K-means clustering groups samples based on features.

```{r}
clusters <- kmeans(dataset[,2:4], centers = 2)

clusters$cluster
```

```{r}
dataset$cluster <- clusters$cluster

dataset
```

### Exercises

13. Count samples per cluster.
14. Compare clusters with responder status.

---

## Section 8 — Exploring Clusters

```{r}
table(dataset$cluster)
```

### Exercises

15. Determine which cluster has more responders.
16. Compare mean GC content per cluster.

---

## Section 9 — Model Comparison

```{r}
model1 <- lm(length ~ gc_content, data = dataset)
model2 <- lm(length ~ gc_content + motif_count, data = dataset)

anova(model1, model2)
```

### Exercises

17. Compare models.
18. Determine which explains more variance.

---

## Section 10 — Small Analysis Pipeline

```{r}
analysis <- dataset |>
  transform(mean_feature = (gc_content + motif_count)/2)

analysis
```

### Exercises

19. Add a new feature column.
20. Use it in a regression model.

---

## Final Exercise

21–25. Build a small analysis workflow that:

- fits a logistic regression model
- generates predicted probabilities
- runs PCA
- performs clustering
- compares clusters with responder status

---

## Skills Gained After Notebook 5

You should now understand:

- linear models (`lm`)
- logistic regression (`glm`)
- interpreting model summaries
- dimensionality reduction with PCA
- clustering with k-means
- building simple statistical analysis workflows

---

## Completion

You have now completed the **five-part R programming and analysis series**:

1. R reboot
2. structured data manipulation
3. tidyverse workflows
4. visualization with ggplot2
5. statistics and modeling

Together these notebooks provide a practical foundation for **data analysis and statistical workflows in computational biology**.
